---
title: "Consensus Approach"
subtitle: "Initial Benchmarking"
author: "Marc Schwering"
output:
  html_document:
    number_sections: true
    toc: true
    toc_depth: 2
    fig_width: 10
    fig_height: 7
    theme: cosmo
---





# Introduction

scDD using pooled cells and KS test (scDD_pool_ks), 
and edgeR using summed up batches (edgeR_sum)
seemed interesting in the initial benchmarking.
scDD_pool_ks is quite powerful and nicely holds FDR control if the batch effect
is not too strong.
edgeR_sum is only good at detecting differential means, but holds FDR control
even under strong batch effects.
Here, I want to investigate both methods.

```{r, message=FALSE, results='hide'}
library(data.table) # database-like tables
library(ggplot2) # plotting
library(scater) # scRNA-seq
library(scDD) # call DD
library(SummarizedExperiment) # using scDD
library(edgeR) # call DD
library(BiocParallel) # parallel
```

I have to recreate the results of these methods again.
Since scDD completely loses FDR control in the strong batch effects scenarios,
only the weak batch effects scenarios would be of interest.

```{r}
paths <- expand.grid(c("strongDD", "weakDD"), "weakBatch", 
                     c("sim1", "sim2", "sim3"))
paths <- apply(paths, 1, function(x) {
  sprintf("%s_%s%s%s", x[1], x[2], .Platform$file.sep, x[3])
})
scenarios <- expand.grid(
  DiffDistr = c("strongDD", "weakDD"),
  Simulation = c("sim1", "sim2", "sim3")
)
```

From the results summary I know that simulation 3 of the strongDD, weakBatch
scenario failed.

```{r}
paths <- paths[-5]
scenarios <- scenarios[-5, ]
```








***




# Reproduce Results

I need the genewise results of each method.
Therefore I need to reproduce the results.

```{r}
edgeR_results <- function(sce, grpID = "group", 
                          btchID = "batch", smplID = "sample") {
  batches <- as.factor(sce[[btchID]])
  l <- lapply(levels(batches), function(i) {
    rowSums(counts(sce)[, which(batches == i)])
  })
  m <- do.call("cbind", l)
  groups <- rep(c("Group1", "Group2"), each = 3)
  dge <- DGEList(counts = m, group = groups)
  dge <- calcNormFactors(dge)
  design <- model.matrix(~ groups)
  dge <- estimateDisp(dge, design)
  fit <- glmFit(dge, design)
  lrt <- glmLRT(fit, coef = 2)
  
  # get results
  res <- topTags(lrt, n = Inf)$table
  res$gene <- as.numeric(
    gsub("([A-Za-z]+)([0-9]+)", "\\2", rownames(res))
  )
  res$rank <- 1:nrow(res)
  res <- res[, -3:-1]
  names(res)[1] <- "pVal"
  res <- as.data.table(res)
  res <- res[order(res$gene), ]
  res$label <- fData(sce)$label
  return(res)
}

scDD_results <- function(sce, grpID = "group", 
                          btchID = "batch", smplID = "sample") {
  groups <- sce[[grpID]]
  m <- exprs(sce)
  m[m < 0] <- 0
  m <- 2^m - 1
  se <- SummarizedExperiment(
    assays = list("NormCounts" = m),
    colData = data.frame(condition = groups)
  )
  se <- scDD(se, permutations = 0, testZeroes = FALSE,
             prior_param = list(alpha = 0.01, mu0 = 0, s0 = 0.01, 
                                a0 = 0.01, b0 = 0.01))
  # get results
  res <- results(se)
  pVals <- res$nonzero.pvalue
  fdrs <- res$nonzero.pvalue.adj
  genes <- as.numeric(
    gsub("([A-Za-z]+)([0-9]+)", "\\2", as.character(res$gene))
  )
  res <- data.table(
    pVal = pVals, 
    FDR = fdrs, 
    gene = genes, 
    rank = rank(pVals)
  )
  res <- res[order(res$gene), ]
  res$label <- fData(sce)$label
  return(res)
}
```

This function will reproduce the results tables for these 2 methods.

```{r, results='hide'}
FUN <- function(path, FedgeR, FscDD) {
  sce <- readRDS(file.path(path, "intermediate_data", "normed.rds"))
  edgeR <- FedgeR(sce)
  scDD <- FscDD(sce)
  edgeR[, method := "edgeR_sum"]
  scDD[, method := "scDD_pool_ks"]
  return(rbind(edgeR, scDD))
}
```

```{r, results='hide'}
l <- bplapply(paths, FUN, FedgeR = edgeR_results, FscDD = scDD_results)
for (i in seq_along(l)) {
  l[[i]]$DiffDistr <- scenarios$DiffDistr[i]
  l[[i]]$Simulation <- scenarios$Simulation[i]
}
res <- rbindlist(l)
saveRDS(res, "results.rds")
```




***









# Overlap

scDD and edgeR work completely differently and so I wondered what kind of genes
they each call.
Below, the relative overlap of genes identified by both methods at different
FDR thresholds is calculated.
That is the size of intersection set divided by the size of the union set
of genes identified by both methods.

```{r}
ths <- seq(.01, 1, .01)
ol <- function(x, y) {
  length(intersect(na.omit(x), na.omit(y))) / 
    length(union(na.omit(x), na.omit(y)))
}
l <- lapply(ths, function(th) {
  dt <- res[FDR <= th, gene, by = .(method, Simulation, DiffDistr)]
  dt <- dcast(dt, Simulation + DiffDistr + gene~ method, value.var = "gene")
  dt[, .(Overlap = ol(edgeR_sum, scDD_pool_ks)), by = .(Simulation, DiffDistr)]
})
for (i in seq_along(l)) l[[i]]$FDR <- ths[i]
overlap <- rbindlist(l)
```

The relative overlap for all simulations with a weak batch effect 
(where both edgeR and scDD hold FDR control)
is shown below.

```{r, results='hide'}
overlap[, group := paste0(DiffDistr, Simulation)]
ggplot(overlap, aes(color = DiffDistr, group = group)) +
  geom_line(aes(x = FDR, y = Overlap)) +
  theme_bw()
```

This plot confirms the assumption that both methods identify rather different
genes.






***










# edgeR and scDD Ensemble

Naturally, one would like to leverage the strengths of both methods.
The easiest way to do this would be to take the union of identified 
genes by both methods.
I will call this ensemble method *consensus approach* here.
To simulate this approach for each gene and simulation the lower FDR from
edgeR and scDD is taken as result.

```{r, results='hide'}
cons <- dcast(res, DiffDistr + Simulation + gene + label ~ method, 
              value.var = "FDR")
cons[, consensus := pmin(edgeR_sum, scDD_pool_ks, na.rm = TRUE)]
cons <- melt(cons, 1:4, 5:7, "method", "FDR")
```

Ranks are calculated again.

```{r}
cons <- cons[, .(gene = gene, label = label, FDR = FDR, rank = rank(FDR)), 
             by = .(DiffDistr, Simulation, method)]
```

Now recall, fallout, true and predicted FDRs have to be calculated again
for a spectrum of different ranks.
The function below does that from ranks, labels and FDRs.

```{r}
getROC <- function(ranks, labels, FDRs, n = 1000, incl = NULL) {
  stopifnot(length(ranks) == length(labels) & length(ranks) == length(FDRs))
  
  # prepare table
  if (!is.null(incl)) {
    idx <- which(labels %in% incl)
    ranks <- ranks[idx]
    labels <- labels[idx]
    FDRs <- FDRs[idx]
  }
  dt <- data.table(rank = ranks, label = labels, FDR = FDRs)
  steps <- seq.int(2, nrow(dt), length.out = n)
  dt <- dt[order(rank)]
  
  # make confusion matrix
  cm <- matrix(NA, n, 4, dimnames = list(steps, c("TP", "FP", "TN", "FN")))
  for( s in seq_along(steps) ){
    res <- dt[, .(TP = sum(rank < steps[s] & label != "random"), 
                    FP = sum(rank < steps[s] & label == "random"), 
                    TN = sum(rank >= steps[s] & label == "random"), 
                    FN = sum(rank >= steps[s] & label != "random"))]
    cm[s, ] <- as.numeric(res[1, ])
  }
  
  # get ROC and FDRs
  dt <- data.table(
    rec = cm[, 1] / (cm[, 1] + cm[, 4]),
    fal = cm[, 2] / (cm[, 2] + cm[, 3]),
    trueFDR = cm[, 2] / (cm[, 2] + cm[, 1]),
    predFDR = dt[steps - 1, FDR]
  )
  return(dt)
}
```




## Prediction Power

Now prediction power and FDR control can be calculated like before.

**Full**

```{r}
sims <- cons[, 1, by = .(DiffDistr, Simulation, method)]
l <- lapply(seq_len(nrow(sims)), function(i) {
  dt <- cons[DiffDistr == sims[i, DiffDistr] & method == sims[i, method] & 
               Simulation == sims[i, Simulation]]
  test <- getROC(dt$rank, dt$label, dt$FDR, incl = NULL)
  test$DiffDistr <- sims[i, DiffDistr]
  test$Simulation <- sims[i, Simulation]
  test$method <- sims[i, method]
  test
})
ROCs <- rbindlist(l)
```

For all types of differential distribution ROC curves are shown below.

```{r, fig.height=5}
ROCs[, group := paste0(method, Simulation)]
ggplot(ROCs, aes(group = group)) +
  geom_path(aes(x = fal, y = rec, color = method)) +
  geom_abline(slope = 1, color = "gray", linetype = 2) +
  facet_grid(~ DiffDistr) +
  theme_bw() +
  ggtitle("ROC -- All types")
```

The consensus method is substantially more powerful than both other methods.

```{r, fig.height=5}
dt <- ROCs[, .(AUC = flux::auc(fal, rec)), 
         by = .(method, DiffDistr, Simulation)]
ggplot(dt, aes(color = method)) +
  geom_jitter(aes(x = method, y = AUC), width = .1, height = 0) +
  facet_grid(~ DiffDistr) +
  theme_bw() +
  ggtitle("AUC -- All types")
```


**D.Shape**

```{r}
l <- lapply(seq_len(nrow(sims)), function(i) {
  dt <- cons[DiffDistr == sims[i, DiffDistr] & method == sims[i, method] & 
               Simulation == sims[i, Simulation]]
  test <- getROC(dt$rank, dt$label, dt$FDR, incl = c("random", "shape"))
  test$DiffDistr <- sims[i, DiffDistr]
  test$Simulation <- sims[i, Simulation]
  test$method <- sims[i, method]
  test
})
ROCshape <- rbindlist(l)
```

Of particular interest for me is how well the method can identify changes other
than differential means.
Therefore, D. Shape is shown here separately.
The consensus method might suffer from edgeR's predictions in the case of
differential shape.

```{r, fig.height=5}
ROCshape[, group := paste0(method, Simulation)]
ggplot(ROCshape, aes(group = group)) +
  geom_path(aes(x = fal, y = rec, color = method)) +
  geom_abline(slope = 1, color = "gray", linetype = 2) +
  facet_grid(~ DiffDistr) +
  theme_bw() +
  ggtitle("ROC -- D.Shape only")
```

The plot shows that this does not seem to be the case.

```{r, fig.height=5}
dt <- ROCshape[, .(AUC = flux::auc(fal, rec)), 
         by = .(method, DiffDistr, Simulation)]
ggplot(dt, aes(color = method)) +
  geom_jitter(aes(x = method, y = AUC), width = .1, height = 0) +
  facet_grid(~ DiffDistr) +
  theme_bw() +
  ggtitle("AUC -- D.Shape only")
```


## FDR Control

Another important issue is FDR control.
Intuitively the consensus approach would inherit FDR control from
the more promiscuous single method.

```{r, fig.height=5}
ROCs[, group := paste0(method, Simulation)]
ggplot(ROCs, aes(group = group)) +
  geom_path(aes(x = trueFDR, y = predFDR, color = method)) +
  geom_abline(slope = 1, color = "gray", linetype = 2) +
  facet_grid(~ DiffDistr) +
  theme_bw() +
  ggtitle("FDR Control")
```

The difference between true and predicted FDRs for predicted FDRs at 
1\%, 5\%, and 10\% are shown below.
Same simulations are indicated by a gray line.

```{r, fig.height=5}
ROCs[, D_FDR := trueFDR - predFDR]
FDRs <- ROCs[, .(FDR.01 = D_FDR[which.min(abs(predFDR - 0.01))],
             FDR.05 = D_FDR[which.min(abs(predFDR - 0.05))],
             FDR.10 = D_FDR[which.min(abs(predFDR - 0.10))]), 
        by = .(method, DiffDistr, Simulation)]
ggplot(FDRs, aes(x = method, y = FDR.01)) +
  geom_line(aes(group = Simulation), color = "gray") +
  geom_point(aes(color = method)) +
  facet_grid(~ DiffDistr) +
  theme_bw() +
  ggtitle("FDR Control at 1% Predicted FDR")
ggplot(FDRs, aes(x = method, y = FDR.05)) +
  geom_line(aes(group = Simulation), color = "gray") +
  geom_point(aes(color = method)) +
  facet_grid(~ DiffDistr) +
  theme_bw() +
  ggtitle("FDR Control at 5% Predicted FDR")
ggplot(FDRs, aes(x = method, y = FDR.10)) +
  geom_line(aes(group = Simulation), color = "gray") +
  geom_point(aes(color = method)) +
  facet_grid(~ DiffDistr) +
  theme_bw() +
  ggtitle("FDR Control at 10% Predicted FDR")
```

FDR control seems to increase additively for the consensus method.
This would make sense.
The plot below shows the sum of edgeR and scDD loss of FDR control
vs the consensus FDR control.

```{r, fig.height=5}
sums <- FDRs[method != "consensus", 
             .(FDR.01 = sum(FDR.01), FDR.05 = sum(FDR.05), FDR.10 = sum(FDR.10)), 
             by = .(DiffDistr, Simulation)]
sums <- melt(sums, 1:2, 3:5, "pred", "sum_edgeRscDD")
cons <- FDRs[method == "consensus", .(FDR.01, FDR.05, FDR.10),
             by = .(DiffDistr, Simulation)]
cons <- melt(cons, 1:2, 3:5, "pred", "consensus")
sums$consensus <- cons$consensus
ggplot(sums) +
  geom_abline(slope = 1, color = "gray", linetype = 2) +
  geom_point(aes(x = sum_edgeRscDD, y = consensus, color = Simulation)) +
  facet_grid(~ DiffDistr) +
  theme_bw()
```











***












# Session Info

```{r}
sessionInfo()
```



