---
title: "Estimate Batch Effects"
author: "Marc Schwering"
output:
  html_document:
    number_sections: true
    toc: true
    toc_depth: 2
    fig_width: 10
    fig_height: 7
    theme: cosmo
---



```{r optsSet, echo=FALSE, results="hide"}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE , dpi = 400)
options(digits = 2)
```

# Load

Packages and functions are loaded, the sce object is laoded, 
and variables are set.

```{r, warning=FALSE, message=FALSE}
grpID <- "group"
btchID <- "batch"
smplID <- "sample"

library(scran) # normalization
library(scater) # single cell expression set
library(ggplot2) # plotting
library(data.table) # data base like tables
library(kBET) # batch effect estimation
library(limma) # estimate robust inter-batch correlations

sce <- readRDS("intermediate_data/normed.rds")
```

Define some variables.
Here, batches 1 to 3 belong to group 1 and the others to group 2.

```{r}
batches <- unique(sce[[btchID]])
groups <- unique(sce[[grpID]])
btchIDX <- rep(1:2, each = 3)
```






# Cell Number

We thought of important measures to capture the batch effect of a dataset.
One important measure might be the average number of cells per batch.
When pooling cells from several batches, pooled cells are falsly assumed as
indepentend measurements by the statistical test.
Since this is not true, the severness of this assumption should somehow be
related to the actual number of cells per batch.

```{r}
(n_hat <- as.vector(ncol(sce) / length(batches)))
```


# Within Batch Variability

The batch effect is probably ralted to the relation of variability 
within batches and the variability between batches.
Here, we calculated the within batch variability as the 
average gene distribution coefficient of variation within each batch, 
averaged over all batches.
We do that on the library size normalized values before log transformation.

```{r}
cv <- function(x) sd(x) / mean(x)
meanCVs <- numeric(length(batches))
for (i in seq_along(batches)) {
  x <- apply(exp(exprs(sce)[, sce[[btchID]] == batches[i]]), 1, cv)
  meanCVs <- mean(x)
}
(sig_hat_within <- mean(meanCVs))
```


# Between Batch Variability

For between batch variability we calculate the coefficient of variation
between the average genewise expressions of batches within groups.
These values are then averaged over genes and groups.
We do that on the library size normalized values before log transformation.

```{r}
CVs <- numeric(2)
for (grp in 1:2) {
  grpBatches <- batches[btchIDX == grp]  
  l <- lapply(grpBatches, function(batch) {
    rowMeans(exp(exprs(sce)[, sce[[btchID]] == batch]))
  })
  CVs[grp] <- mean(apply(do.call(cbind, l), 1, cv))
}
(sig_hat_between <- mean(CVs))
```




# kBET

Using [kBET](https://github.com/theislab/kBET) Pearson's $\chi^2$ tests
are performed to compare the mixture of batches of the overall dataset
with the mixture of batches in local neighbourhoods.
Local neighbourhoods are defined by a k-nearest-neighbour kernel.
Here, we use the ratio of local to global expected means
as a measure of the batch effect.
We will also use non-transformed values here, as the package does not 
specify otherwise.

```{r}
res <- kBET(t(exp(exprs(sce))), sce[[btchID]], plot = FALSE)$summary
(kBET_rate <- res$kBET.observed[1] - res$kBET.expected[1])
```

The significance will also be saved.

```{r}
(kBET_pVal <- res$kBET.signif[1])
```


# LMM

Fit a linear mixed model with a random effect for bacthes and a fixed effect
for each gene by REML.
Here, we use the average estimated inter-batch correlation as a measure for
the batch effect.
The average is the trimmed mean of the individual correlations on the 
atanh-transformed scale.

```{r}
corrs <- duplicateCorrelation(
  exprs(sce), 
  design = model.matrix(~ 1 + sce[[grpID]]), 
  block = sce[[btchID]]
)
(batch_corrs <- corrs$consensus.correlation)
```






# Save

Final save.

```{r}
estimates <- list(
  n_hat = n_hat,
  sig_hat_within = sig_hat_within,
  sig_hat_between = sig_hat_between,
  kBET_rate = kBET_rate,
  kBET_pVal = kBET_pVal,
  batch_corrs = batch_corrs
)
saveRDS(estimates, "results/estimates.rds")
```




***
