---
title: "Patel 2014"
subtitle: "Comparison of Stemness Genesets"
author: "Marc Schwering"
output:
  html_document:
    number_sections: true
    toc: true
    toc_depth: 2
    fig_width: 10
    fig_height: 7
    theme: cosmo
---





# Preparation

## Load

Libraries are loaded.

```{r, message=FALSE, results='hide'}
library(scater)
library(data.table)
library(ggplot2)
library(SummarizedExperiment)
library(edgeR)
library(scDD)
library(DT)
library(ggthemes)
```

Load data and set variables.

```{r}
grpID <- "Group"
btchID <- "Replicate"
smplID <- "Cell"
sce <- readRDS("intermediate_data/sce.rds")
sce <- sce[fData(sce)$label == "endo", ]
```

Load genesets.

```{r}
setPatel <- as.character(read.csv("data/StemnessSignature.csv")$Gene)
setSchwering <- as.character(read.csv("results/Stemness_geneset.csv")$Gene)
str(setPatel)
str(setSchwering)
genelist <- readRDS("intermediate_data/Stemness_genelist.rds")
```



















***
















# Genewise Comparison

Here, I will take a look at which genes were identified by each
filter and why.
The below function is for plotting Venn diagrams.

```{r}
VennDiag <- function(l, labs = names(l), cols = c("#4285f4", "#db4437", "#0f9d58", "#f4b400")) {
  cols <- rep(cols, length.out = length(l))
  names(l) <- labs
  grid::grid.newpage()
  futile.logger::flog.threshold(futile.logger::ERROR, name = "VennDiagramLogger")
  grid::grid.draw(VennDiagram::venn.diagram(
    l, filename = NULL, na = "remove", cex = 1.5, lty = 1, lwd = 1, cat.cex = 1,
    fill = cols, 
    main.cex = 1.5, main.fontfamily = 2, cat.fontfamily = 2
  ))  
}
```

## Overview

The below Venn diagram should give an overview of both gene sets.

```{r}
VennDiag(list(Patel = setPatel, Schwering = setSchwering))
(setBoth <- intersect(setPatel, setSchwering))
```

Even though I set quite strict FDR threshold, my set is much larger
then that from Patel et al.
Overlap is very small.

## Intersect

Below are plots and a table of the genes which appear in both sets.

```{r}
idxBoth <- which(fData(sce)$hgncSymb %in% setBoth)
df <- data.frame(
  exprs = as.vector(t(exprs(sce[idxBoth, ]))),
  gene = rep(fData(sce)$hgncSymb[idxBoth], each = ncol(sce)),
  Group = rep(sce[[grpID]], length(idxBoth))
)
ggplot(df, aes(color = Group, fill = Group, x = Group, y = exprs)) +
  geom_jitter(width = .1, size = .5) +
  geom_violin(alpha = .3) +
  facet_wrap(~ gene) +
  labs(y = "log10 Expression") +
  theme_hc()
DT::datatable(genelist[HGNC %in% setBoth], rownames = FALSE, 
              options = list(pageLength = 10)) %>%
  formatRound(c(3, 4, 7))
```

I am surprised that FDFT1 is in the Patel set.
Patel filtered the genes to have at least 1.5 groupwise fold change.
It is only present in my set because scDD identified the distributions
as having a different number of modes in both comparisons.



## Difference Patel

Below is the set of genes which were identified by Patel but not by me.
In Patel et al they were identified by a minimum per group fold change and
a higher minimum mean fold change.

```{r}
setX <- setdiff(setPatel, setSchwering)
DT::datatable(genelist[HGNC %in% setX], rownames = FALSE, 
              options = list(pageLength = 20)) %>%
  formatRound(c(3, 4, 7))
```

Some genes show contradicting fold changes, and some have a quite low
mean fold change.
Since in Patel more groups (also the confounded groups without replicates)
were considered it can be that the mean fold change was higher for these
genes.
The main reason why most of these genes were not identified in my
filter is the FDR.
I set a quite strict FDR.
Some of these genes appear borderline significant, but not below 0.1\% FDR.
The few that are below 0.1\% FDR do not have a high absolute mean fold change.

```{r}
idxX <- which(fData(sce)$hgncSymb %in% setX)
idxs <- sample(idxX, 12)
df <- data.frame(
  exprs = as.vector(t(exprs(sce[idxs, ]))),
  gene = rep(fData(sce)$hgncSymb[idxs], each = ncol(sce)),
  Group = rep(sce[[grpID]], length(idxs))
)
ggplot(df, aes(color = Group, fill = Group, x = Group, y = exprs)) +
  geom_jitter(width = .1, size = .5) +
  geom_violin(alpha = .3) +
  facet_wrap(~ gene) +
  labs(y = "log10 Expression") +
  theme_hc()
```


## Difference Schwering

Below is a table of the genes which were identified by my filter but not by
Patel et al's.

```{r}
setX <- setdiff(setSchwering, setPatel)
DT::datatable(genelist[HGNC %in% setX & Filter, ], rownames = FALSE, 
              options = list(pageLength = 20)) %>%
  formatRound(c(3, 4, 7))
```

There are many genes with high group wise fold changes and 
high mean fold changes.
These should have also been identified by the filter from Patel.
I can only assume that the reason why they were not identified is, that
the other group included by Patel (the confounded groups without replicates)
lowered the mean fold change too much.

```{r}
idxX <- which(fData(sce)$hgncSymb %in% setX)
idxs <- sample(idxX, 12)
df <- data.frame(
  exprs = as.vector(t(exprs(sce[idxs, ]))),
  gene = rep(fData(sce)$hgncSymb[idxs], each = ncol(sce)),
  Group = rep(sce[[grpID]], length(idxs))
)
ggplot(df, aes(color = Group, fill = Group, x = Group, y = exprs)) +
  geom_jitter(width = .1, size = .5) +
  geom_violin(alpha = .3) +
  facet_wrap(~ gene) +
  labs(y = "log Expression") +
  theme_hc()
```












***




















# Terms

## Gene Set Enrichment Analysis

Some tools for GSEA.

```{r}
source("R/doGSEA.R")
load("data/ENSG2category.rda")
```

GO terms for cell cycle, biological process and molecular function
will be tested for enrichment using Fisher's exact test based on the
set of genes identified by each filter.
Furthermore, some html files are written, in order to browse terms.

**Patel**

```{r, warning=FALSE}
idx <- which(genelist$HGNC %in% setPatel)
gseaPatel <- doGSEA(genelist$ensID[idx], genelist$ensID, ENSG2category)
plotGSEA(gseaPatel, nrOfSets = 25, main = "Enriched GO Terms in Patel Set")
DT::datatable(tableGSEA(gseaPatel), rownames = FALSE, options = list(pageLength = 20))
```

**Schwering**

```{r, warning=FALSE}
idx <- which(genelist$HGNC %in% setSchwering)
gseaSchwering <- doGSEA(genelist$ensID[idx], genelist$ensID, ENSG2category)
plotGSEA(gseaSchwering, nrOfSets = 25, main = "Enriched GO Terms in Schwering Set")
DT::datatable(tableGSEA(gseaSchwering), rownames = FALSE, options = list(pageLength = 20))
```



## Comparison

Here, I want to compare the relevance significance of terms in both gene sets.

```{r}
TermsPatel <- setDT(gseaPatel$T)
TermsSchwering <- setDT(gseaSchwering$T)
TermsPatel <- TermsPatel[order(ID)]
TermsSchwering <- TermsSchwering[order(ID)]
```

Both GSEAs are summarized in a table.

```{r, results='hide'}
Terms <- data.table(
  Term = TermsPatel$ID,
  Name = TermsPatel$Name,
  Patel = TermsPatel$padj,
  Schwering = TermsSchwering$padj,
  Ann = TermsPatel$Ann,
  Source = TermsPatel$source
)
```

The venn diagram below shows the overlap of enriched GO terms
using a 10\% FDR threshold.

```{r}
VennDiag(list(Patel = Terms[Patel < 0.1, Term], 
              Schwering = Terms[Schwering < 0.1, Term]))
int <- intersect(Terms[Patel < 0.1, Name], Terms[Schwering < 0.1, Name])
DT::datatable(Terms[Name %in% int], rownames = FALSE, options = list(pageLength = 20))
```

Obviously as with the genes there is little overlap.
Most terms are very general.
The only specific terms are related to cell migration, localization,
and motility.

**Agreement**

To see how well terms of the GSEA agree from both gene sets,
I will compute the relative overlap of terms identified at various
FDR thresholds.

```{r}
overlap <- function(Terms, ths = seq(.01, 1, .01)) {
  ol <- function(x, y) {
    length(intersect(na.omit(x), na.omit(y))) / 
      length(union(na.omit(x), na.omit(y)))
  }
  l <- lapply(ths, function(th) {
    X <- Terms[Patel <= th, Term]
    Y <- Terms[Schwering <= th, Term]
    data.table(Overlap = ol(X, Y), FDR = th)
  })
  rbindlist(l)
}
```

Below is the area chart to show the relative overlap at various FDR's.
As comparison, the relative overlap in case of random terms is shown as well.

```{r}
ol <- overlap(Terms)
ol$Rep <- 0
ol$Sets <- "PatelSchwering"
l <- lapply(1:100, function(i) {
  T1 <- Terms
  T1$Patel <- runif(nrow(T1))
  T1$Schwering <- runif(nrow(T1))
  T1 <- overlap(T1)
  T1$Rep <- i
  T1
})
olr <- rbindlist(l)
olr$Sets <- "Random"
dt <- rbind(olr, ol)
ggplot(dt, aes(x = FDR, y = Overlap)) +
  geom_path(aes(group = Rep, color = Sets)) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 0.5)) +
  scale_color_hc() +
  theme_hc()
```

For FDR's up to around 5\% the overlap is similar to a random overlap.
After that, the overlap is substantially lower than random.
Below is another visualization to show the disagreement between
GSEA's from both datasets.

```{r}
t <- table(Terms$Source)
n <- names(t[t > 500])
Terms[!Source %in% n, Source := "Other"]
Terms$Source <- factor(Terms$Source, levels(as.factor(Terms$Source))[c(1:4, 6, 5)])
```

```{r}
ggplot(Terms, aes(size = Ann, color = Source)) +
  geom_point(aes(x = -log10(Patel), y = -log10(Schwering))) +
  labs(y = "-log10 FDR Schwering", x = "-log10 FDR Patel") +
  scale_size_continuous(name = "Annotations") +
  theme_bw()
Terms[, .(PCorr = cor(Patel, Schwering)), by = Source]
```




















***



# Session


```{r}
sessionInfo()
```



